---
title: "Hyeseon RNAseq - DESeq2"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,
  fig.height = 5
  )
```

## Read in count files

```{r, message=FALSE}
library(tidyverse)


# list all count files -- these include boostrapped counts
count_files <- list.files("../results",
                          recursive = TRUE,
                          pattern = ".ReadsPerGene.out.tab$",
                          full.names = TRUE)

# remove bootstrapped counts from count files
count_files <- count_files[!str_detect(count_files, "boot")]

# read count files into a single dataframe
count_df <- vroom::vroom(count_files,
                         delim = "\t",
                         col_names = c("gene", "unstranded", "forward", "reverse"),
                         id = "fpath")

head(count_df)
```

## Clean up count files

- remove N_ rows 
- select only unstranded counts
- extract sample_names from files paths

```{r}
library(tools)
library(magrittr)


count_df %<>% 
  filter(!str_detect(gene, "^N_")) %>% 
  mutate(sample_name = basename(file_path_sans_ext(fpath)),
         sample_name = str_remove(sample_name, ".ReadsPerGene.out")) %>% 
  select(gene, sample_name, unstranded) %>% 
  filter(unstranded > 0)

head(count_df)
```

## Coerce count_df to a count matrix

```{r}
count_mat <- count_df %>% 
  pivot_wider(names_from = sample_name, 
              values_from = unstranded,
              values_fill = 0) %>% 
  column_to_rownames(var = "gene")

head(count_mat)
```

## Create colData for DEseq2

```{r}
col_data <- colnames(count_mat) %>% 
  enframe(name = "row_id", value = "sample_name") %>% 
  separate(sample_name, 
           into = c("age", "region", "passage"),
           sep = "_",
           remove = FALSE) %>% 
  mutate(group = factor(str_c(age, region, sep = ".")),
         passage = factor(passage)) %>% 
  select(sample_name, group, passage) %>% 
  column_to_rownames(var = "sample_name")

head(col_data)
```

## Read data into deseq2 object and run DE pipeline

In the design matrix below we treat passage as a covariate and we will treat
the concatenation of age and organ each as a single group. This will allow
us to correct for the effect of passage (which is appearing as a strong 
batch effect) if we wish to treat each passage as a biological replicate.

```{r message=FALSE, warning=FALSE}
library(DESeq2)


dds <- DESeqDataSetFromMatrix(countData = count_mat,
                              colData = col_data,
                              design = ~0 + group + passage)

dds <- DESeq(dds)

# view the groups that will be used in the contrasts
resultsNames(dds)
```

## DE results

### Set significance and FC levels

```{r}
sig <- 0.1
fc <- 1.5
```

### Old vs. Young

The contrast below indicates averaging the expression for "groupO.A" and 
"groupO.T" and comparing it to the expression of "groupY.A".

```{r}
# load helper functions for plotting
library(coriell)  # devtools::install_github("coriell-research/coriell")


# extract DE results from DESeq object
ovy <- results(dds, 
              contrast = c(0.5, 0.5, -1, 0),
              lfcThreshold = log2(fc),
              alpha = sig,
              altHypothesis = "greaterAbs")

# convert to dataframe
ovy_df <- ovy %>% 
  as_tibble(rownames = "gene") %>% 
  mutate(padj = replace_na(padj, 1))

# show volcano plot
plot_volcano(ovy_df, 
             x = log2FoldChange,
             y = padj,
             fdr = sig,
             lfc = log2(fc)) +
  ggtitle("Old vs Young") +
  ggsave("../results/figures/ovy-volcano.deseq.png",
         width = 8,
         height = 5)

# show MD plot
plot_md(ovy_df,
        x = log10(baseMean),
        y = log2FoldChange,
        sig_col = padj,
        fdr = sig,
        lfc = log2(fc)) +
  ggtitle("Old vs Young") +
  ggsave("../results/figures/ovy-md.deseq.png",
         width = 8,
         height = 5)
```

### Thoracic vs. Aortic

```{r}
# extract DE results from DESeq object
tva <- results(dds, 
              contrast = c(-0.5, 1, -0.5, 0),
              lfcThreshold = log2(fc),
              alpha = sig,
              altHypothesis = "greaterAbs")

# convert to dataframe
tva_df <- tva %>% 
  as_tibble(rownames = "gene") %>% 
  mutate(padj = replace_na(padj, 1))

# show volcano plot
plot_volcano(tva_df, 
             x = log2FoldChange,
             y = padj,
             fdr = sig,
             lfc = log2(fc)) +
  ggtitle("Thoracic vs. Aortic") +
  ggsave("../results/figures/tva-volcano.deseq.png",
         width = 8,
         height = 5)

# show MD plot
plot_md(tva_df,
        x = log10(baseMean),
        y = log2FoldChange,
        sig_col = padj,
        fdr = sig,
        lfc = log2(fc)) +
  ggtitle("Thoracic vs. Aortic") +
  ggsave("../results/figures/tva-md.deseq.png",
         width = 8,
         height = 5)
```

### Old Aortic vs. Young Aortic

```{r}
# extract DE results from DESeq object
oA_v_yA <- results(dds, 
                  contrast = c(1, 0, -1, 0),
                  lfcThreshold = log2(fc),
                  alpha = sig,
                  altHypothesis = "greaterAbs")

# convert to dataframe
oA_v_yA_df <- oA_v_yA %>% 
  as_tibble(rownames = "gene") %>% 
  mutate(padj = replace_na(padj, 1))

# show volcano plot
plot_volcano(oA_v_yA_df, 
             x = log2FoldChange,
             y = padj,
             fdr = sig,
             lfc = log2(fc)) +
  ggtitle("Old Aortic vs. Young Aortic") +
  ggsave("../results/figures/oA_v_yA-volcano.deseq.png",
         width = 8,
         height = 5)

# show MD plot
plot_md(oA_v_yA_df,
        x = log10(baseMean),
        y = log2FoldChange,
        sig_col = padj,
        fdr = sig,
        lfc = log2(fc)) +
  ggtitle("Old Aortic vs. Young Aortic")+
  ggsave("../results/figures/oA_v_yA-md.deseq.png",
         width = 8,
         height = 5)
```

### Old Thoracic vs Old Aortic

```{r}
# extract DE results from DESeq object
oT_v_OA <- results(dds, 
                   contrast = c(-1, 1, 0, 0),
                   lfcThreshold = log2(fc),
                   alpha = sig,
                   altHypothesis = "greaterAbs")

# convert to dataframe
oT_v_oA_df <- oT_v_OA %>% 
  as_tibble(rownames = "gene") %>% 
  mutate(padj = replace_na(padj, 1))

# show volcano plot
plot_volcano(oT_v_oA_df, 
             x = log2FoldChange,
             y = padj,
             fdr = sig,
             lfc = log2(fc)) +
  ggtitle("Old Thoracic vs. Old Aortic") +
  ggsave("../results/figures/oT_v_oA-volcano.deseq.png",
         width = 8,
         height = 5)

# show MD plot
plot_md(oT_v_oA_df,
        x = log10(baseMean),
        y = log2FoldChange,
        sig_col = padj,
        fdr = sig,
        lfc = log2(fc)) +
  ggtitle("Old Thoracic vs. Old Aortic") +
  ggsave("../results/figures/oT_v_oA-md.deseq.png",
         width = 8,
         height = 5)
```

## Combine all results into a single dataframe and write out

```{r}
res_list <- list("deseq.old_vs_young" = ovy_df,
                 "deseq.thoracic_vs_aortic" = tva_df,
                 "deseq.oldArotic_vs_youngAortic" = oA_v_yA_df,
                 "deseq.oldThoracic_vs_oldAortic" = oT_v_oA_df)

bind_rows(res_list, .id = "contrast") %>%
  write_tsv("../results/data-files/deseq2-dge.tsv")
```

